---
title: "Correlation and Simple Linear Regression Demo"
author: "Professor Christopher J. Schmank"
date: "2024-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(pacman)
p_load(dplyr,tidyr,psych,skimr,ggplot2,gridExtra,jmv,mvtnorm,gapminder,moderndive,forcats,broom,janitor)
p_loaded()
```

## Lets load in the data for this demo

```{r}
load(url("http://www.openintro.org/stat/data/evals.RData"))
evals <- evals %>%
  select(score, bty_avg, age)
```

## Exploratory Data Analysis

1. Observe raw data
1. Compute descriptive statistics
1. Visualize data

## Observe raw data

```{r}
head(evals)
glimpse(evals)
#View(evals)
```

## Compute descriptive statistics

```{r}
# Can use any of these but my preference is the `describe()` function
describe(evals)

# These alternatives give you many descriptive stats but nothing about normality
skim(evals)
summary(evals)
skim_without_charts(evals)

# Don't forget correlations!
cor(evals$bty_avg,evals$score)
corrMatrix(evals)
```

## Visualize the data

```{r}
ggplot(data = evals, 
       mapping = aes(x = score)) +
  geom_histogram(bins = 10, 
                 color = "seagreen", 
                 fill="black")

ggplot(data = evals, 
       mapping = aes(x = bty_avg)) +
  geom_histogram(bins = 10, 
                 color = "seagreen", 
                 fill="black")

ggplot(data = evals, 
       mapping = aes(x = age)) +
  geom_histogram(bins = 10, 
                 color = "seagreen", 
                 fill="black")
```

```{r}
ggplot(data = evals, 
       mapping = aes(y = score)) +
  geom_boxplot()

ggplot(data = evals, 
       mapping = aes(y = bty_avg)) +
  geom_boxplot()

ggplot(data = evals, 
       mapping = aes(y = age)) +
  geom_boxplot()
```

```{r}
ggplot(evals, 
       aes(x = bty_avg, 
           y = score)) +
  geom_point() +
  labs(x = "Beauty Score", 
       y = "Teaching Score",
       title = "Relationship of Teaching and Beauty Score") +
  geom_smooth(method = "lm")

ggplot(evals, 
       aes(x = age, 
           y = score)) +
  geom_point() +
  labs(x = "Age", 
       y = "Teaching Score",
       title = "Relationship of Teaching and Age") +
  geom_smooth(method = "lm")
```


## Now lets run our simple linear regression (`lm()` workflow + `broom` package)
```{r}
score_model <- lm(score ~ bty_avg, data = evals)

# Get regression table using `tidy()`:
tidy(score_model, conf.int=TRUE)

# Get model fit table using `glance()`:
glance(score_model)
```

## Alternatively we can use `linReg()` from the `jmv` package
```{r}
jmv::linReg(evals,
            dep=score,
            covs=bty_avg,
            blocks=list(list('bty_avg')),
            anova=TRUE,
            stdEst=TRUE,
            ci=TRUE,
            ciStdEst=TRUE)
```

## Lastly we should conduct a residual analysis

```{r}

regression_points <- get_regression_points(score_model)

# Scatterplot of Explanatory Variable along x-axis and Residuals along y-axis
ggplot(regression_points, 
       aes(x = bty_avg, y = residual)) +
  geom_point() +
  labs(x = "Beauty Score", y = "Residual") +
  geom_hline(yintercept = 0, col = "blue", size = 1)

# Additionally we want to look at the distribution of residuals
ggplot(regression_points, 
       aes(x = residual)) +
  geom_histogram(bins = 10, color = "forestgreen") +
  labs(x = "Residual")

describe(regression_points$residual)
```

