---
title: "Continuous Data Analysis Demo --- The `t` Tests"
author: "Professor Christopher J. Schmank"
date: "2024-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
# install.packages("datarium")
# install.packages("rstatix")
# install.packages("ggpubr")

library(jmv)
library(psych)
library(tidyverse)
library(infer)
library(datarium)
library(rstatix)
library(ggpubr)
```

## One-Sample `t`-test

One numerical variable

**Research Question 1**: Is the average weight of mice different from 25g? What about less than 25g?

```{r}
# Load and inspect demo data set(s)

# Load `mice` data from `datarium` package
data(mice, package="datarium")
head(mice)
```

**Hypothesis Statement**: 2-tailed (Non-directional)

$H_0$: Null Hypothesis --- Average weight of sampled mice is not different from 25 g, $\mu_{sample} = \mu_{population} = \text{25 grams}$

$H_a$: Alternate Hypothesis --- Average weight of sampled mice is different from 25 g, $\mu_{sample} \neq \mu_{population} \neq \text{25 grams}$

**Hypothesis Statement**: 1-tailed (Directional)

$H_0$: Null Hypothesis --- Average weight of sampled mice is not less than 25 g, $\mu_{sample} \geq \mu_{population} \geq \text{25 grams}$

$H_a$: Alternate Hypothesis --- Average weight of sampled mice is less than 25 g, $\mu_{sample} < \mu_{population} < \text{25 grams}$

---

**Research Question 2**: Is the average hours of work different from 40 hours per week? What about greater than 40 hours per week?

```{r}
set.seed(13)
# Load `gss` data from `infer` package
data(gss, package="infer") 
gss <- gss %>% 
  select(hours) %>% 
  rep_sample_n(size = 100, replace = TRUE)
head(gss)
```

**Hypothesis Statement**: 2-tailed (Non-directional)

$H_0$: Null Hypothesis --- Average hours of work per week is not different from 40, $\mu_{sample} = \mu_{population} = \text{40 hours per week}$

$H_a$: Alternate Hypothesis --- Average hours of work per week is different from 40, $\mu_{sample} \neq \mu_{population} \neq \text{40 hours per week}$

**Hypothesis Statement**: 1-tailed (Directional)

$H_0$: Null Hypothesis --- Average hours of work per week is not greater than from 40, $\mu_{sample} \leq \mu_{population} \leq \text{40 hours per week}$

$H_a$: Alternate Hypothesis --- Average hours of work per week is different from 40, $\mu_{sample} > \mu_{population} > \text{40 hours per week}$

## Descriptive Statistics (Do not forget to visualize!!)

```{r}
#Descriptive statistics for `weight` variable
describe(mice$weight)

#Descriptive statistics for `hours` variable
describe(gss$hours)

```

## `jmv` One-Sample `t`-test `ttestOneS()`

**Research Question 1**
```{r}
mice %>% 
ttestOneS(vars=weight,
          testValue = 25,
          hypothesis = 'dt',
          norm = TRUE,
          ci = TRUE,
          effectSize = TRUE,
          ciES = TRUE,
          desc = TRUE)

mice %>% 
ttestOneS(vars=weight,
          testValue = 25,
          hypothesis = 'lt',
          norm = TRUE,
          ci = TRUE,
          effectSize = TRUE,
          ciES = TRUE,
          desc = TRUE)
```

## Publishable Plot | Using `rstatix` and `ggpubr` packages

```{r}
stat_test <- mice %>% rstatix::t_test(weight ~ 1, mu = 25, alternative="two.sided") 
stat_test

# `alternative =` can be set to `greater` or `less` as well

# Density Plot

ggdensity(mice, 
          x = "weight", 
          rug = TRUE, 
          fill = "lightgray") +
  scale_x_continuous(limits = c(15, 27)) +
  stat_central_tendency(type = "mean", color = "red", linetype = "dashed") +
  geom_vline(xintercept = 25, color = "blue", linetype = "dashed") + 
  labs(subtitle = get_test_label(stat_test,  detailed = TRUE))
```

**Research Question 2**
```{r}
gss %>% 
ttestOneS(vars=hours,
          testValue = 40,
          hypothesis = 'dt',
          norm = TRUE,
          ci = TRUE,
          effectSize = TRUE,
          ciES = TRUE,
          desc = TRUE)

gss %>% 
ttestOneS(vars=hours,
          testValue = 40,
          hypothesis = 'gt',
          norm = TRUE,
          ci = TRUE,
          effectSize = TRUE,
          ciES = TRUE,
          desc = TRUE)

# Violations of the Shapiro-Wilks test can set `wilcoxon = TRUE` argument
# gss %>% 
# ttestOneS(vars=hours,
#           testValue = 40,
#           hypothesis = 'dt',
#           norm = TRUE,
#           ci = TRUE,
#           effectSize = TRUE,
#           ciES = TRUE,
#           desc = TRUE,
#           wilcoxon = TRUE)
# 
# gss %>% 
# ttestOneS(vars=hours,
#           testValue = 40,
#           hypothesis = 'gt',
#           norm = TRUE,
#           ci = TRUE,
#           effectSize = TRUE,
#           ciES = TRUE,
#           desc = TRUE,
#           wilcoxon = TRUE)
```

```{r}
stat_test <- gss %>% rstatix::t_test(hours ~ 1, mu = 40, alternative = "greater")
stat_test

# Density Plot

ggdensity(gss, 
          x = "hours", 
          rug = TRUE, 
          fill = "lightgray") +
  scale_x_continuous(limits = c(0, 90)) +
  stat_central_tendency(type = "mean", color = "red", linetype = "dashed") +
  geom_vline(xintercept = 40, color = "blue", linetype = "dashed") + 
  labs(subtitle = get_test_label(stat_test,  detailed = TRUE))
```

## APA Style Results Section

**Statistical Interpretation**: 

A one-sample *t*-test was performed to evaluate whether there was a difference between the weight in a sample of mice (*n* = 10) compared to the general population of mice ($\mu = 25$). The mean weight in the sample of mice (M = 20.14, SD = 1.90) was significantly different than the average weight of mice in the general population, *t*(9) = -8.11, *p* < .001, Cohen's *d* = -2.56, 95% CI [-3.87, -1.23].


---

## Paired-Sample (Repeated-Measure) `t`-test

Numerical variable collected twice OR `within-subject` DV

**Research Question 1**: Does the weight of 10 mice differ from `Before` to `After` following a prescribed treatment? Does the treatment lead to increased weight from `Before` to `After` for the 10 mice?

```{r}
# Load and inspect demo data set(s)

# Load `mice2` data from `datarium` package
data(mice2, package="datarium")
head(mice2)

# Transform into long data: 
# gather the before and after values in the same column
mice2_long <- mice2 %>%
  gather(key = "group", value = "weight", before, after)
head(mice2_long)
```

**Hypothesis Statement**: 2-tailed (Non-directional)

$H_0$: Null Hypothesis --- Average weight of sampled mice is not different due to treatment (i.e., from `Before` to `After`), $\mu_{Before} = \mu_{After} \text{ OR } \mu_{Before - After} = 0$

$H_a$: Alternate Hypothesis --- Average weight of sampled mice is different due to treatment (i.e., from `Before` to `After`), $\mu_{Before} \neq \mu_{After}  \text{ OR } \mu_{Before - After} \neq 0$

**Hypothesis Statement**: 1-tailed (Directional)

$H_0$: Null Hypothesis --- Average weight of sampled mice is not greater following treatment (i.e., from `Before` to `After`), $\mu_{Before} \geq \mu_{After} \text{ OR } \mu_{Before - After} \geq 0$

$H_a$: Alternate Hypothesis --- Average weight of sampled mice is greater following treatment (i.e., from `Before` to `After`), $\mu_{Before} < \mu_{After} \text{ OR } \mu_{Before - After} < 0$

---

**Research Question 2**:

```{r}
# Load and inspect demo data set(s)

# Load `depression` data from `datarium` package
data(depression, package = "datarium")
depression<- depression %>%
  select(id,t0,t3)
head(depression)

# Transform into long data: 
# gather the before and after values in the same column
dep_long <- depression %>%
  gather(key = "time_period", value = "depression", t0, t3)
head(dep_long)
```

**Hypothesis Statement**: 2-tailed (Non-directional)

$H_0$: Null Hypothesis --- Average depression level of sampled patients is not different from `Baseline` to `Follow-Up`, $\mu_{Baseline} = \mu_{Follow-Up} \text{ OR } \mu_{Baseline - Follow-Up} = 0$

$H_a$: Alternate Hypothesis --- Average depression level of sampled patients is different from `Baseline` to `Follow-Up`, $\mu_{Baseline} \neq \mu_{Follow-Up}  \text{ OR } \mu_{Baseline - Follow-Up} \neq 0$

**Hypothesis Statement**: 1-tailed (Directional)

$H_0$: Null Hypothesis --- Average depression level of sampled patients is not lower from `Baseline` to `Follow-Up`, $\mu_{Baseline} \geq \mu_{Follow-Up} \text{OR } \mu_{Baseline - Follow-Up} \geq 0$

$H_a$: Alternate Hypothesis --- Average depression level of sampled patients is lower from `Baseline` to `Follow-Up`, $\mu_{Baseline} < \mu_{Follow-Up} \text{OR } \mu_{Baseline - Follow-Up} < 0$

## Descriptive Statistics (Do not forget to visualize!!)

```{r}
# Change `group` into a factor
mice2_long$group<-factor(mice2_long$group)

# Use the `describeBy` function to isolate `weight`
describeBy(mice2_long[,3],group=mice2_long$group,mat=TRUE)

# Change `time_period` into a factor
dep_long$time_period<-factor(dep_long$time_period)

# Use the `describeBy` function to isolate `depression`
describeBy(dep_long[,3],group=dep_long$time_period,mat=TRUE)
```

## `jmv` Paired-Sample `t`-test `ttestPS()`

**Research Question 1**
```{r}
mice2 %>% 
ttestPS(pairs=list(list(i1 = "before", i2 = "after")),
        hypothesis = 'different',
        norm = TRUE,
        ci = TRUE,
        effectSize = TRUE,
        ciES = TRUE,
        desc = TRUE)

mice2 %>% 
ttestPS(pairs=list(list(i1 = "before", i2 = "after")),
        hypothesis = 'twoGreater',
        norm = TRUE,
        ci = TRUE,
        effectSize = TRUE,
        ciES = TRUE,
        desc = TRUE)
```

## Publishable Plot | Using `rstatix` and `ggpubr` packages

```{r}
stat_test <- mice2_long  %>%
  rstatix::t_test(weight ~ group, ref.group = "before", paired = TRUE, alternative = "two.sided") %>%
  add_significance()
stat_test

# `alternative =` can be set to `greater` or `less` as well

# Side-by-Side Boxplot Comparison

# Create a box plot
bxp <- ggboxplot(mice2_long, 
                x = "group", 
                y = "weight",
                order = c("before", "after"),
                ylab = "Weight", xlab = "Groups")

# Add p-value and significance levels
stat_test <- stat_test %>% add_xy_position(x = "group")

bxp + 
  stat_pvalue_manual(stat_test, tip.length = .03) +
  labs(subtitle = get_test_label(stat_test, detailed= TRUE))
```

**Research Quesiton 2**
```{r}
depression %>% 
ttestPS(pairs=list(list(i1 = "t0", i2 = "t3")),
        hypothesis = 'different',
        norm = TRUE,
        ci = TRUE,
        effectSize = TRUE,
        ciES = TRUE,
        desc = TRUE)

depression %>% 
ttestPS(pairs=list(list(i1 = "t0", i2 = "t3")),
        hypothesis = 'oneGreater',
        norm = TRUE,
        ci = TRUE,
        effectSize = TRUE,
        ciES = TRUE,
        desc = TRUE)

```

```{r}
stat_test <- dep_long  %>%
  rstatix::t_test(depression ~ time_period, paired = TRUE, ref.group = "t0", alternative = "greater") %>%
  add_significance()
stat_test

# `alternative =` can be set to `greater` or `less` as well

# Side-by-Side Boxplot Comparison

# Create a box plot
bxp <- ggboxplot(dep_long, 
                x = "time_period", 
                y = "depression",
                order = c("t0", "t3"),
                ylab = "Depression Level", xlab = "Time Period")

# Add p-value and significance levels
stat_test <- stat_test %>% add_xy_position(x = "time_period")

bxp + 
  stat_pvalue_manual(stat_test, tip.length = 0.01) +
  labs(subtitle = get_test_label(stat_test, detailed= TRUE))
```

## APA Style Results Section

**Statistical Interpretation**: 

A paired-sample *t*-test was conducted to determine whether there was a difference between the depression level among our sample from baseline (`t0`) to a 6 month follow-up (`t3`). The results indicated that depression level at baseline (*M* = 292.38, *SD* = 73.87) was statistically significantly higher than depression level at 6 month follow-up (*M* = 151.92, *SD* = 78.02), *t*(23) = 6.69, *p* < .001, Cohen's *d* = 1.37, 95% CI [.798, 1.918].

---

## Independent-Samples (Between-Subject) `t`-test

Numerical DV; Mutually Exclusive, Binary Grouping Variable

**Research Question 1**: Does the average stress level of 60 participants differ due to a randomly assigned stress treatment ($n_{perGroup} = 30$)? Was the average stress level of the `treatment group` lower than the average stress level of the `non-treatment group`?

```{r}
# Load and inspect demo data set(s)

# Load `mice2` data from `datarium` package
data(stress, package="datarium")
stress <- stress %>%
  select(id,score,treatment)
head(stress)

```

**Hypothesis Statement**: 2-tailed (Non-directional)

$H_0$: Null Hypothesis --- Average stress level was not different between treatment groups (i.e., from `Treatment` to `Non-treatment`), $\mu_{Treatment} = \mu_{Non-Treatment} \text{ OR } \mu_{Treatment } - \mu_{Non-Treatment} = 0$

$H_a$: Alternate Hypothesis --- Average stress level was different between treatment groups (i.e., from `Treatment` to `Non-treatment`), $\mu_{Treatment} \neq \mu_{Non-Treatment} \text{ OR } \mu_{Treatment } - \mu_{Non-Treatment} \neq 0$

**Hypothesis Statement**: 1-tailed (Directional)

$H_0$: Null Hypothesis --- Average stress level was not lower between the `treatment` and `non-treatment` groups, $\mu_{Treatment} \geq \mu_{Non-Treatment} \text{ OR } \mu_{Treatment } - \mu_{Non-Treatment} \geq 0$

$H_a$: Alternate Hypothesis --- Average stress level was lower between the `treatment` and `non-treatment` groups, $\mu_{Treatment} < \mu_{Non-Treatment} \text{ OR } \mu_{Treatment } - \mu_{Non-Treatment} < 0$

---

**Research Question 2**: Does the average increase in hours of sleep across 20 participants differ due to a randomly assigned treatment ($n_{perGroup} = 10$)? Was the average increase in hours of sleep of the `medication group` greater than the average increase in hours of sleep of the `Control group`?

```{r}
# Load and inspect demo data set(s)

# Load `depression` data from `datarium` package
data(sleep, package = "datasets")
head(sleep)
```

**Hypothesis Statement**: 2-tailed (Non-directional)

$H_0$: Null Hypothesis --- Average increases in hours of sleep was not different between treatment groups (i.e., from `Medication` to `Control`), $\mu_{Medication} = \mu_{Control} \text{ OR } \mu_{Medication } - \mu_{Control} = 0$

$H_a$: Alternate Hypothesis --- Average increases in hours of sleep was different between treatment groups (i.e., from `Medication` to `Control`), $\mu_{Medication} \neq \mu_{Control} \text{ OR } \mu_{Medication } - \mu_{Control} \neq 0$

**Hypothesis Statement**: 1-tailed (Directional)

$H_0$: Null Hypothesis --- Average increases in hours of sleep was not greater from `Medication` to `Control Treatment Groups`, $\mu_{Medication} \leq \mu_{Control} \text{ OR } \mu_{Medication } - \mu_{Control} \leq 0$

$H_a$: Alternate Hypothesis --- Average increases in hours of sleep was greater from `Medication` to `Control Treatment Groups`, $\mu_{Medication} > \mu_{Control} \text{ OR } \mu_{Medication } - \mu_{Control} > 0$

## Descriptive Statistics (Do not forget to visualize!!)

```{r}
# Use the `describeBy` function
describeBy(stress$score,group=stress$treatment,mat=TRUE)

describeBy(sleep$extra,group=sleep$group,mat=TRUE)
```

## `jmv` Independent-Samples `t`-test `ttestIS()`

**Research Question 1**
```{r}
stress %>% 
ttestIS(vars= "score",
        group = "treatment",
        hypothesis = 'different',
        norm = TRUE,
        eqv = TRUE,
        ci = TRUE,
        effectSize = TRUE,
        ciES = TRUE,
        desc = TRUE)

stress %>% 
ttestIS(vars= "score",
        group = "treatment",
        hypothesis = 'twoGreater',
        norm = TRUE,
        eqv = TRUE,
        ci = TRUE,
        effectSize = TRUE,
        ciES = TRUE,
        desc = TRUE)

# Violations of the Levenes test can set `welchs = TRUE` argument
# stress %>% 
# ttestIS(vars= "score",
#         group = "treatment",
#         hypothesis = 'different',
#         norm = TRUE,
#         eqv = TRUE,
#         ci = TRUE,
#         effectSize = TRUE,
#         ciES = TRUE,
#         desc = TRUE,
#         welchs = TRUE)
# 
# stress %>% 
# ttestIS(vars= "score",
#         group = "treatment",
#         hypothesis = 'twoGreater',
#         norm = TRUE,
#         eqv = TRUE,
#         ci = TRUE,
#         effectSize = TRUE,
#         ciES = TRUE,
#         desc = TRUE,
#         welchs = TRUE)
```

## Publishable Plot | Using `rstatix` and `ggpubr` packages

```{r}
stat_test <- stress  %>%
  rstatix::t_test(score ~ treatment, ref.group = "yes", alternative = "two.sided", var.equal = TRUE) %>%
  add_significance()
stat_test

# `alternative =` can be set to `greater` or `less` as well

# Side-by-Side Boxplot Comparison

# Create a box plot
bxp <- ggboxplot(stress, 
                x = "treatment", 
                y = "score",
                order = c("yes", "no"),
                ylab = "Stress Level", xlab = "Treatment Condition")

# Add p-value and significance levels
stat_test <- stat_test %>% add_xy_position(x = "treatment")

bxp + 
  stat_pvalue_manual(stat_test, tip.length = 0) +
  labs(subtitle = get_test_label(stat_test, detailed= TRUE))
```

**Research Question 2**
```{r}
sleep %>% 
ttestIS(vars= "extra",
        group = "group",
        hypothesis = 'different',
        norm = TRUE,
        eqv = TRUE,
        ci = TRUE,
        effectSize = TRUE,
        ciES = TRUE,
        desc = TRUE)

sleep %>% 
ttestIS(vars= "extra",
        group = "group",
        hypothesis = 'twoGreater',
        norm = TRUE,
        eqv = TRUE,
        ci = TRUE,
        effectSize = TRUE,
        ciES = TRUE,
        desc = TRUE)

```

```{r}
stat_test <- sleep  %>%
  rstatix::t_test(extra ~ group, ref.group = "1", alternative = "less", var.equal = TRUE) %>%
  add_significance()
stat_test

# `alternative =` can be set to `greater` or `less` as well

# Side-by-Side Boxplot Comparison

# Create a box plot
bxp <- ggboxplot(sleep, 
                x = "group", 
                y = "extra",
                order = c("1", "2"),
                ylab = "Extra Hours of Sleep", xlab = "Treatment Group")

# Add p-value and significance levels
stat_test <- stat_test %>% add_xy_position(x = "group")

bxp + 
  stat_pvalue_manual(stat_test, tip.length = 0) +
  labs(subtitle = get_test_label(stat_test, detailed= TRUE))
```

## APA Style Results Section

**Statistical Interpretation**: 

An independent samples *t*-test was conducted to determine whether there was a difference between the extra hours of sleep obtained by participants who received the experimental medication and participants who received the placebo. The results indicated that participants who received the medication (*M* = 2.33, *SD* = 2.00) had significantly greater levels of extra hours of sleep obtained compared to participants who received the placebo (*M* = 0.75, *SD* = 1.79), *t*(18) = -1.86, *p* = .079, Cohen's *d* = -0.83, 95% CI [-1.739, 0.096].