---
title: "Infer Pipeline --- Continuous Data Analysis Tutorials"
author: "Professor Christopher J. Schmank"
date: "2024-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, messages=FALSE,warning=FALSE}
library(tidyverse)
library(infer)
```

## `infer` One-Sample `t`-test

One numerical variable (standardized mean `t`)

```{r}
# Generate `observed` t-statistic
t_bar <- gss %>%
  specify(response = hours) %>%
  hypothesize(null = "point", mu = 40) %>%
  calculate(stat = "t")
```

```{r}
# Generate distribution assuming null (H_0)
null_dist <- gss %>%
  specify(response = hours) %>%
  hypothesize(null = "point", mu = 40) %>%
  generate(reps = 1000) %>%
  calculate(stat = "t")

# Generate theoretical t-distribution
null_dist_theory <- gss %>%
  specify(response = hours)  %>%
  assume("t")
```

```{r}
# Visualizing the inferential test (2-tailed)
visualize(null_dist, method = "both") +
  shade_p_value(obs_stat = t_bar, direction = "two-sided")

# Visualizing the inferential test (1-tailed)
visualize(null_dist, method = "both") +
  shade_p_value(obs_stat = t_bar, direction = "greater")

# Extract p-value of t-test (2-tailed)
null_dist %>%
  get_p_value(obs_stat = t_bar, direction = "two-sided")

# Extract p-value of t-test (1-tailed)
null_dist %>%
  get_p_value(obs_stat = t_bar, direction = "greater")

# Options for direction =  c("two-sided", "greater", or "less")
```

```{r}
set.seed(13)

# Confidence Interval around `t`-statistic

# Generate bootstrapped Null Distribution
boot_dist <- gss %>%
   specify(response = hours) %>%
   hypothesize(null = "point", mu = 40) %>%
   generate(reps = 1000, type = "bootstrap") %>%
   calculate(stat = "t")

standard_error_ci <- boot_dist %>%
  get_ci(type = "se", point_estimate = t_bar)

visualize(boot_dist) +
  shade_confidence_interval(endpoints = standard_error_ci)
```



## `infer` Paired Sample `t`-test

One numerical variable (Paired/Repeated Measure) --- Mean of Differences

```{r}
set.seed(13)

gss_paired <- gss %>%
   mutate(
      hours_previous = hours + 5 - rpois(nrow(.), 4.8),
      diff = hours - hours_previous
   )

gss_paired %>%
   select(hours, hours_previous, diff)
```


```{r}
# Generate `observed` statistic(s)
x_tilde <- gss_paired %>%
  specify(response = diff) %>%
  calculate(stat = "mean")

t_diff <- gss_paired %>% 
  specify(response = diff) %>% 
  hypothesize(null = "point", mu = 0) %>%
  calculate(stat = "t")
```

```{r}
# Generate distribution assuming null (H_0)
null_dist <- gss_paired %>%
  specify(response = diff) %>%
  hypothesize(null = "paired independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "mean")
```

```{r}
# Visualizing the inferential test (2-tailed)
visualize(null_dist) +
  shade_p_value(obs_stat = x_tilde, direction = "two-sided")

# Visualizing the inferential test (1-tailed)
visualize(null_dist) +
  shade_p_value(obs_stat = x_tilde, direction = "less")

# Extract p-value of t-test (2-tailed)
null_dist %>%
  get_p_value(obs_stat = x_tilde, direction = "two-sided")

# Extract p-value of t-test (1-tailed)
null_dist %>%
  get_p_value(obs_stat = x_tilde, direction = "less")

```

```{r}
set.seed(13)

# Confidence Interval around mean of differences

# Generate bootstrapped Null Distribution
boot_dist <- gss_paired %>%
   specify(response = diff) %>%
   hypothesize(null = "point", mu = 0) %>%
   generate(reps = 1000, type = "bootstrap") %>%
   calculate(stat = "mean")

standard_error_ci <- boot_dist %>%
  get_ci(type = "se", point_estimate = x_tilde)

visualize(boot_dist) +
  shade_confidence_interval(endpoints = standard_error_ci)
```

```{r}
# Generate distribution assuming null (H_0)
null_dist <- gss_paired %>%
  specify(response = diff)  %>%
  hypothesize(null = "point", mu = 0) %>% 
  generate(reps = 1000) %>% 
  calculate(stat = "t")
```

```{r}
# Visualizing the inferential test (2-tailed)
visualize(null_dist) +
  shade_p_value(obs_stat = t_diff, direction = "two-sided")

# Visualizing the inferential test (1-tailed)
visualize(null_dist) +
  shade_p_value(obs_stat = t_diff, direction = "less")

# Extract p-value of t-test (2-tailed)
null_dist %>%
  get_p_value(obs_stat = t_diff, direction = "two-sided")

# Extract p-value of t-test (1-tailed)
null_dist %>%
  get_p_value(obs_stat = t_diff, direction = "less")

```


```{r}
set.seed(13)

# Confidence Interval around mean of differences

# Generate bootstrapped Null Distribution
boot_dist <- gss_paired %>%
   specify(response = diff) %>%
   hypothesize(null = "point", mu = 0) %>%
   generate(reps = 1000, type = "bootstrap") %>%
   calculate(stat = "t")

standard_error_ci <- boot_dist %>%
  get_ci(type = "se", point_estimate = t_diff)

visualize(boot_dist) +
  shade_confidence_interval(endpoints = standard_error_ci)
```


## `infer` Independent Samples `t`-test

One numerical, one categorical (2 levels ONLY) variables --- Standardized difference in means


```{r}
# Generate `observed` statistic
t_hat <- gss %>% 
  specify(age ~ college) %>% 
  hypothesize(null = "independence") %>%
  calculate(stat = "t", order = c("degree", "no degree"))
```

```{r}
# Generate distribution assuming null (H_0)
null_dist <- gss %>%
  specify(age ~ college) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "t", order = c("degree", "no degree"))

# Generate theoretical t-distribution
null_dist_theory <- gss %>%
  specify(age ~ college) %>%
  assume("t")
```

```{r}
# Visualizing the inferential test (2-tailed)
visualize(null_dist, method = "both") +
  shade_p_value(obs_stat = t_hat, direction = "two-sided")

# Visualizing the inferential test (1-tailed)
visualize(null_dist, method = "both") +
  shade_p_value(obs_stat = t_hat, direction = "greater")

# Extract p-value of t-test (2-tailed)
null_dist %>%
  get_p_value(obs_stat = t_hat, direction = "two-sided")

# Extract p-value of t-test (1-tailed)
null_dist %>%
  get_p_value(obs_stat = t_hat, direction = "greater")

```

```{r}
set.seed(13)

# Confidence Interval around mean of differences

# Generate bootstrapped Null Distribution
boot_dist <- gss %>%
   specify(hours ~ college) %>%
   hypothesize(null = "independence") %>% 
   generate(reps = 1000, type = "bootstrap") %>%
   calculate(stat = "t", order = c("degree", "no degree"))

standard_error_ci <- boot_dist %>%
  get_ci(type = "se", point_estimate = t_hat)

visualize(boot_dist) +
  shade_confidence_interval(endpoints = standard_error_ci)
```


## `infer` One-way ANOVA `F`-test

One numerical, one categorical (>2 levels) variables


```{r, message=FALSE}
# Generate `observed` statistic
F_hat <- gss %>% 
  specify(age ~ partyid) %>%
  calculate(stat = "F")
```

```{r message=FALSE}
# Generate distribution assuming null (H_0)
null_dist <- gss %>%
   specify(age ~ partyid) %>%
   hypothesize(null = "independence") %>%
   generate(reps = 1000, type = "permute") %>%
   calculate(stat = "F")

# Generate theoretical t-distribution
null_dist_theory <- gss %>%
   specify(age ~ partyid) %>%
   hypothesize(null = "independence") %>%
   assume(distribution = "F")
```

```{r}
# Visualizing the inferential test (1-tailed)
visualize(null_dist, method = "both") +
  shade_p_value(obs_stat = F_hat, direction = "greater")

# Extract p-value of t-test (1-tailed)
null_dist %>%
  get_p_value(obs_stat = F_hat, direction = "greater")

```