---
title: "PSY 5790: Applied Univariate and Bivariate/Bivariable Statistics"
subtitle: "Continuous Data Analysis Demo 2 --- ANOVA"
author: "Christopher J. Schmank, PhD"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, message=FALSE,warning=FALSE}
# install.packages("datarium")
# install.packages("rstatix")
# install.packages("ggpubr")

library(jmv)
library(psych)
library(tidyverse)
library(infer)
library(datarium)
library(rstatix)
library(ggpubr)
```

## Repeated-Measures (Within-Subject) ANOVA

Single DV collected multiple times from Ss

**Research Question 1**: Are the self-esteem means different across the three time points for our participants?

```{r}
# Load and inspect demo data set(s)

# Load `mice` data from `datarium` package
data(selfesteem, package="datarium")
head(selfesteem)
```

**Hypothesis Statement**: 2-tailed (Non-directional)

$H_0$: Null Hypothesis --- Self-esteem means are not different across time, $\mu_{Time1} = \mu_{Time2} = \mu_{Time3}$

$H_a$: Alternate Hypothesis --- At least one mean ($\mu$) differs from another

## Descriptive Statistics (Do not forget to visualize!!)

First, we will change the format from wide to long!

```{r}
selfesteemB <- selfesteem %>%
  gather(key = "time", value = "selfest", t1, t2, t3) %>%
  convert_as_factor(id, time)
head(selfesteemB, 3)

#Descriptive statistics for `self-esteem` variable across `time`
describeBy(selfesteemB$selfest, group=selfesteemB$time, mat=TRUE)

```

## `jmv` Repeated Measures ANOVA

### Assumption Checks

```{r}
# Normality of DV across levels of IV tests

selfesteemB %>%
  group_by(time) %>%
  shapiro_test(selfest)

ggqqplot(selfesteemB, 
         "selfest", 
         facet.by = "time")
```


```{r}
# Sphericity check for when 3 + RM values used!

selfesteem %>%
  anovaRM(rm = list(
    list(label = 'Timing',
         levels = c('Time1','Time2','Time3'))),
    rmCells = list(
      list(measure = 't1',
           cell = 'Time1'),
      list(measure = 't2',
           cell = 'Time2'),
      list(measure = 't3',
           cell = 'Time3')),
    spherTests = TRUE) #This is the only information we really need to interpret AT THIS POINT!!

```

# Running the RM ANOVA: `jmv`


```{r}
selfesteem %>%
  anovaRM(rm = list(list(
    label = 'Time',
    levels = c('Time 1','Time 2','Time 3'))),
         rmCells = list(
           list(
             measure = 't1',
             cell = 'Time 1'),
           list(
             measure = 't2',
             cell = 'Time 2'),
           list(
             measure = 't3',
             cell = 'Time 3')),
    effectSize = 'eta',
    spherCorr = list(c('none','GG','HF')),
    spherTests = TRUE,
    postHoc = 'Time',
    postHocCorr = 'bonf',
    depLabel = "Self-Esteem",
    emMeans = list(list("Time")),
    emmPlots = TRUE,
    groupSumm = TRUE)

```

## Publishable Plot | Using `rstatix` and `ggpubr` packages
```{r}
# Generate pairwise comparison table with xy positioning

pwc <- selfesteemB %>%
  pairwise_t_test(selfest ~ time, 
                  paired = TRUE,
                  p.adjust.method = "bonferroni") %>%
  add_xy_position(x = "time")

pwc
```
```{r}
# Generate publishable plot from long data

bxp <- ggboxplot(selfesteemB, 
                 x = "time", 
                 y = "selfest") +
  labs(x = "Time",
       y = "Self-Esteem Score") +
  stat_pvalue_manual(pwc, hide.ns = TRUE)
bxp
```


## APA Style Results Section

**Statistical Interpretation**: 

A within-subject one-way ANOVA was performed to determine the relationship between self-esteem scores and timing. The mean and standard deviations are presented in Table 1. The ANOVA was statistically significant at the .05 alpha level, *F*(2, 18) = 55.47, *p* \< .001, $\eta^2$ = .83. A post hoc test was conducted using the Bonferroni Correction that indicated that the mean self-esteem level at Baseline were statistically significantly lower than self-esteem levels at Time 1 (*p* = .002) and at Time 2 (*p* \< .001). Furthermore the mean self-esteem level at Time 1 was statistically significantly lower than self-esteem levels at Time 2 (*p* = .003).


## One-Way (Between-Subjects) ANOVA

Single DV collected across **mutually exclusive** groups (3+)

**Research Question 1**: Do different nutrient treatments (e.g., `control`, `calcium`, `nitrogen`) lead to different levels of `plant growth` (i.e., `weight`)? 

```{r}
# Load and inspect demo data set(s)

# Load `mice` data from `datarium` package
data("PlantGrowth")
head(PlantGrowth)
```

**Hypothesis Statement**: 2-tailed (Non-directional)

$H_0$: Null Hypothesis --- Plant growth (i.e., `weight`) means are not different across nutrient treatments, $\mu_{Control} = \mu_{Calcium} = \mu_{Nitrogen}$

$H_a$: Alternate Hypothesis --- At least one mean ($\mu$) differs from another

## Descriptive Statistics (Do not forget to visualize!!)

```{r}
#Descriptive statistics for `weight` variable across `group`
describeBy(PlantGrowth$weight, group=PlantGrowth$group, mat=TRUE)

```


## `jmv` One-way Between-Subjects ANOVA

### Assumption Checks

```{r}
PlantGrowth %>%
  group_by(group) %>%
  shapiro_test(weight)

ggqqplot(PlantGrowth, "weight", facet.by = "group")

PlantGrowth %>% 
  levene_test(weight ~ group)
```

```{r}

# Can also use the `norm = ` and `homo = ` arguments of ANOVA() to assess
# Normality and Homogeneity from a univariate perspective
PlantGrowth %>%
  ANOVA(dep=weight,
        factors = group,
        norm = TRUE,
        homo = TRUE)
```


# Running the ANOVA

```{r}
PlantGrowth %>%
  ANOVA(dep=weight,
        factors = group,
        effectSize = 'eta',
        norm = TRUE,
        homo = TRUE,
        postHoc = weight~group,
        postHocCorr = 'bonf',
        postHocES = 'd',
        postHocEsCi = TRUE,
        emMeans = list(list("group")),
        emmPlots = TRUE)
```

## Publishable Plot | Using `rstatix` and `ggpubr` packages

```{r}
pwcB <- PlantGrowth %>%
  pairwise_t_test(weight ~ group,
                  p.adjust.method = "bonferroni") %>% 
  add_xy_position(x = "group")

pwcB

pwcT <- PlantGrowth %>%
  tukey_hsd(weight ~ group) %>% 
  add_xy_position(x = "group")
pwcT
```


```{r}

ggboxplot(PlantGrowth, 
          x = "group", 
          y = "weight") +
  labs(x = "Plant Growth Treatment Group",
       y = "Plant Weight"
  ) +
  stat_pvalue_manual(pwcB, hide.ns = TRUE, tip.length = 0)

ggboxplot(PlantGrowth, 
          x = "group", 
          y = "weight") +
  labs(x = "Plant Growth Treatment Group",
       y = "Plant Weight"
  ) +
  stat_pvalue_manual(pwcT, hide.ns = FALSE, tip.lenth = 0)

```


## APA Style Results Section

**Statistical Interpretation**: 

A between-subjects one-way ANOVA was performed to determine the relationship between plant growth (i.e., weight) and nutrient treatments (e.g., control, calcium, and nitrogen). The mean and standard deviations are presented in Table 1. The ANOVA was statistically significant at the .05 alpha level, *F*(2, 27) = 4.85, *p* = .016, $\eta^2$ = .264. A post hoc test was conducted using the Bonferroni Correction that indicated that the mean weight of plants treated with Nitrogen were statistically significantly higher than the mean weight of plants treated with Calcium (*p* = .01, Cohen's *d* = -1.388, 95% CI [-2.384, -0.392]). However, the mean weight of plants treated with the control was not statistically significantly different compared to the mean weight of plants treated with Nitrogen (*p* = .58, Cohen's *d* = 0.595, 95% CI [-0.337, 1.528]) or to the mean weight of plants treated with Calcium (*p* = .26, Cohen's *d* = -0.793, 95% CI [-1.736, 0.152]).