---
title: "Psychological Measurement: Descriptive Statistics, Visualizations, and Reliability Tutorial"
author: "Christopher J. Schmank, PhD"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Load Libraries for Tutorial

```{r echo=TRUE,message=FALSE,warning=FALSE}
library(tidyverse)
library(psych)
library(jmv)
```

#### Data Visualization Template `ggplot`

```{r eval=FALSE,echo=TRUE}

ggplot(data = <DATA>, 
       mapping = aes(<MAPPINGS>)) + 
  <GEOM_FUNCTION>()

```

- `data = ` name of data frame/object within R Environment
- `mapping = aes()` argument requires variable name/label from within `data =`
    + What variable(s) do you want to visualize?
    + Different graphics/visualizations require specifying: 
        - `x =` OR `y =` within `aes()` (**univariate**)
        - `x = ` AND `y = ` within `aes()` (**bivariate**/**grouped**)
       
- `geom_function()` is then layered (`+`) into the code to achieve the appropriate visualization:
    + Histograms --- `geom_histogram()`
    + Boxplots --- `geom_boxplot()` 
    + Scatterplots  --- `geom_point()`
    + Linegraphs --- `geom_line()`
    + Barplots --- `geom_bar()` OR `geom_col()`
    
#### Data Visualization | Histograms & Boxplots

**Histogram Examples**:
```{r eval=TRUE,echo=TRUE,warning=FALSE,message=FALSE}

# Univariate Histogram Code
ggplot(data = bfi,
       mapping = aes(x = age)) +
  geom_histogram() 

# Faceted, Bivariate Histogram Code
ggplot(data = bfi,
       mapping = aes(x = age)) +
  geom_histogram() +
  facet_wrap(~ gender, nrow = 2) # Can also use `ncol = 1` for same stacked visualization

```

**Histogram Examples Expanded**
```{r eval=TRUE,echo=TRUE,warning=FALSE,message=FALSE}

# Adding Cutomization Layers to Histogram Visualization
ggplot(data = bfi,
       mapping = aes(x = age)) +
  geom_histogram(bins = 12, color = "goldenrod2") +
  labs(y = "Count", x = "Age of Participant")

bfi$genderF<-factor(bfi$gender, labels = c("Male", "Female"))

ggplot(data = bfi,
       mapping = aes(x = age)) +
  geom_histogram(bins = 12, color = "forestgreen") +
  facet_wrap(~ genderF, nrow = 2) +
  labs(y = "Count", x = "Age of Participant")

```

**Boxplot Examples**
```{r eval=TRUE,echo=TRUE,warning=FALSE,message=FALSE}

# Univariate boxplot
ggplot(data = bfi,
       mapping = aes(y = age)) +
  geom_boxplot() +
  labs(y = "Age of Participant")

#Grouped, Bivariate Boxplot
ggplot(data = bfi,
       mapping = aes(x = genderF, y = age)) +
  geom_boxplot() +
  labs(x = "Gender", y = "Age of Participant")

```


#### Data Visualization | Scatterplots & Linegraphs

**Scatterplot Examples**
```{r eval=TRUE,echo=TRUE,warning=FALSE,message=FALSE}

# Bivariate Scatterplot
ggplot(data = attitude, 
       mapping = aes(x = complaints, y = rating)) + 
  geom_point() +
  labs(x = "How Employee Deals With Complaints Rating", y = "Overall Employee Ratings")

# Bivariate Scatterplot w/ Line of Best-Fit
ggplot(data = attitude, 
       mapping = aes(x = complaints, y = rating)) + 
  geom_point() +
  labs(x = "How Employee Deals With Complaints Rating", y = "Overall Employee Ratings") +
  geom_smooth(method = "lm", se = FALSE)

# Bivariate Scatterplot w/ Line of Best-Fit (+ Standard Error)
ggplot(data = attitude, 
       mapping = aes(x = complaints, y = rating)) + 
  geom_point() +
  labs(x = "How Employee Deals With Complaints Rating", y = "Overall Employee Ratings") +
  geom_smooth(method = "lm", se = TRUE)

```

**Linegraph Example**
```{r eval=TRUE,echo=TRUE,warning=FALSE,message=FALSE}
# Bivariate Linegraph (TIME ACROSS X-AXIS)
ggplot(data = economics, 
       aes(x = date, y = unemploy)) +
  geom_line() + 
  labs(x = "Data Collection Year", y = "Number of Unemployed Persons")

```


#### Describing Data | `psych` Package

```{r eval=FALSE,echo=TRUE}
# Describe entire dataset OR subset of data object
describe(<DATA>)

# Grouped descriptive statistics of entire dataset OR subset of data object
describeBy(<DATA>, 
           group = <GROUPING VARIABLE>,
           mat = TRUE)
```

By default `describe()` and `describeBy()` functions use these arguments:

1. Remove missing data listwise --- `na.rm = TRUE`
1. Calculate skew and kurtosis --- `skew = TRUE`
1. Calculate the range --- `ranges = TRUE`
1. DOES NOT calculate the Interquartile Range --- `IQR = FALSE`

```{r eval=TRUE,echo=TRUE,warning=FALSE,message=FALSE}

describe(bfi[,c(1:5,26,28)], 
         IQR = TRUE)

describeBy(bfi[,c(1:5,28:29)],
           group = bfi$genderF,
           IQR = TRUE,
           mat = TRUE)
```

#### Calculating Correlation | `psych` and `jmv`

```{r eval=FALSE,echo=TRUE}

# `psych` package correlation matrix
cor_data <- corr.test(<DATA>,
                      use = "complete")
print(cor_data, short = FALSE)

# Can call for ONLY correlation matrix OR save as data object
cor_data$r

# `jmv` package correlation matrix and visualization
corrMatrix(<DATA>, 
           vars = vars(<VARIABLE1>, <VARIABLE2>, <VARIABLE3>, ... <VARIABLEk>,),
           ci = TRUE,
           plots = TRUE,
           plotDens=TRUE,
           plotStats=TRUE)
```

**`psych` Correlation Example**
```{r eval=TRUE,echo=TRUE, warning=FALSE,message=FALSE}
# Generate correlation data object 
cor_bfi <- corr.test(bfi[,c(1:5,26,28)],
                     use = "complete")

# Prints all output related to `corr.test()`
print(cor_bfi, short=FALSE)

# Prints only correlation matrix
cor_bfi$r

# Saves correlation matrix as data object for later use
cor_bfi_r <- cor_bfi$r

```

**`jmv` Correlation Example**

```{r eval=TRUE,echo=TRUE, warning=FALSE,message=FALSE}

# Can specify the variables individually
corrMatrix(bfi,
           vars = vars('A1','A2','A3','A4','A5','gender','age'),
           ci=TRUE,
           plots = TRUE,
           plotDens=TRUE,
           plotStats=TRUE)

# Can specify the variables by subsetting data object
corrMatrix(bfi[,c(1:5,26,28)],
           ci=TRUE,
           plots = TRUE,
           plotDens=TRUE,
           plotStats=TRUE)

```

#### Internal Consistency Reliability

**Cronbach's Alpha ($\alpha$)**
```{r eval=FALSE,echo=TRUE}

# Set seed for replicability of bootstrapped CI generation
set.seed(<NUMBER>)

# Create alpha data object for later use
alpha_dat<-alpha(<DATA>,
                 check.keys = TRUE,
                 n.iter = 100,
                 use = "complete.obs")

```

**Alpha Example**
```{r eval=TRUE,echo=TRUE,warning=FALSE,message=FALSE}

# For replication, value of `set.seed()` should remain consistent 
set.seed(13)

# Creates data object that can be printed fully or called specifically
alpha_bfi <- alpha(bfi[1:25],
                   check.keys = TRUE,
                   n.iter = 100,
                   use="complete.obs")

# Call all output related to alpha function
alpha_bfi

# Specific calls available
alpha_bfi$total
alpha_bfi$boot.ci
alpha_bfi$alpha.drop
alpha_bfi$item.stats

# Can use dplyr piping to retrieve Standardized Alpha value
bfi %>%
  dplyr::select(A1:O5) %>%
  drop_na() %>%
  alpha(check.keys = TRUE) %>% # check.keys = TRUE checks for negatively coded items and reverses them!!
  .$total %>% 
  dplyr::select("std.alpha")

```

**McDonald's Omega ($\omega$)**
```{r eval=FALSE,echo=TRUE}

# Create omega data object for later use
omega_dat <- omega(<DATA>,
                   check.keys=TRUE,
                   nfactors = <NUMBER>,
                   fm = "ml",
                   n.iter = 1,
                   use = "complete.obs")

```

**Omega Example**
```{r eval=TRUE,echo=TRUE,warning=FALSE,message=FALSE}

# Creates data object that can be printed fully or called separately
omega_bfi <- omega(bfi[1:25],
                   check.keys=TRUE,
                   nfactors=5,
                   fm="ml",
                   n.iter=1,
                   use = "complete.obs")

# Calls all output related to omega function
omega_bfi

# Specific calls available
omega_bfi$omega.tot
omega_bfi$alpha

# Can use dplyr piping to retrieve Standardized Alpha value
bfi %>%
  dplyr::select(A1:O5) %>%
  drop_na() %>%
  omega(plot=FALSE) %>%
  .$omega.tot
```

**Split-Half Reliability**
```{r eval=FALSE,echo=TRUE}

# Runs split-half reliability estimate on data
splitHalf(<DATA>,
          raw = TRUE,
          check.keys=TRUE,
          use = "complete")
             
```

**Split-Half Example**
```{r eval=TRUE,echo=TRUE,warning=FALSE,message=FALSE}

# Generate split-half reliability output
splitHalf(bfi[1:25],
          raw=TRUE,
          check.keys=TRUE,
          use = "complete")

```